---
title: "Gov 50 Final Project"
author: "Mike Yan"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to inflation in the American economy and presidential approval ratings. I want to especially study the presidential approval ratings all the way from FDR to Joe Biden. Because I'm studying the overall relationship of inflation and approval ratings, collecting data from all the presidents  could reinforce the strength and validity of my results. The results of this study will show whether or not inflation and presidential approval ratings have an inverse relationship and why presidents would often prioritize economy as their primary agenda.

## Proposal

Does inflation negatively impact the presidential approval ratings? In this study, I plan to examine the extent to which economic inflation/deflation in the US affects the incumbent president's job performance rating in the eyes of the American public. I hypothesize that increasing inflation would cause the approval ratings to go down, and a stable economy with low levels of inflation would correspond with a high approval rating. During periods of high inflation, I expect that prices of most products and services in the US would rise. Consumers who have the same levels of income now have to pay more for these products and services, leading them with a lower levels of saving and financial stability. This phenomenon would make consumers disillusioned in the US economy, and this financial distress would translate to political discontent with the president. As a result, I suspect that these consumers, who are also US citizens who can participate in American politics, would report a lower level of approval of the incumbent president. My sample is comprised of all the US presidents from FDR to Joe Biden. My unit of analysis is each president. The explanatory variable of interest is monthly inflation of the US economy from July 1941 to September 2023. These inflation rates are calculated using the Consumer Price Index, which is published monthly by the Bureau of Labor Statistics (BLS) of the U.S. Department of Labor. The outcome variable is the presidential approval rating. It varies from 0-100, which represents the percentage of respondents who express approval/disapproval of the president's performance. For example, if the outcome variable is 54, it means that 54% of the respondents approve the president's overall performance in the past month. The data comes from Gallup's Presidential Approval project, which conducts national polls asking about public approval of the president. The question asked to the respondents is as follows: Do you approve or disapprove of the way [enter President name] is handling his job as President? To investigate inflation's effect on approval rating, I would conduct a regression analysis to test the correlation between inflation and approval rating for each president, and take an average of all the correlation coefficients to obtain a single value. Finally, we would use the significance test to see if the correlation value is statistically significant. If it's statistically significant, then inflation does have an effect on presidential approval rating. If it's not statistically significant, then we cannot find convincing evidence that there's a correlation.

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(zoo)
library(broom)
```

```{r}
inflation <- read_csv("data/Gov_50_Inflation.csv")
approval <- read_csv("data/Gov_50_Pres_Approval.csv")
```
## Approval Rating Example: Approval Rating during Trump's Presidency
```{r}
approval |>
  filter(President == "Trump") |>
  ggplot(aes(x = `Start Date`, y = Approving)) +
  geom_line(linewidth = 1.5, color = "coral2")+
  labs(x = "Date",
       y = "Approval Rating")
```



## Code for Joining the Approval Dataset and the Inflation Dataset
```{r}
inflation <- inflation |>
  select(-Avg) |>
  pivot_longer(
    cols = `Jan`:`Dec`,
    names_to = "month",
    values_to = "inflation") |>
  mutate(year_month = as.yearmon(paste(Year, month), "%Y %b")) |>
  select(-Year, -month)
``` 

```{r}
approval <- approval |>
  mutate(year_month = as.yearmon(`Start Date`,"%Y %b")) |>
  group_by(year_month) |>
  mutate(monthly_approval = mean(Approving))
```

```{r}
approval <- approval |>
  left_join(inflation) |>
  drop_na() |>
  distinct(year_month, .keep_all = TRUE)
```
## Inflation Example: Inflation Rate during Obama's Presidency
```{r}
approval |>
  filter(President == "Obama") |>
  ggplot(aes(x = `Start Date`, y = inflation)) +
  geom_line(linewidth = 1.5, color = "steelblue1")+
  labs(x = "Date",
       y = "Inflation")
```

```{r}
approval2000 <- approval |>
  filter(year_month >= "Jan 2000")
approval2000 |>
  ggplot(aes(x = inflation, y = monthly_approval)) +
  geom_point()
```
## Overall Trend since FDR
```{r}
approval |>
  ggplot() +
  geom_line(aes(x= year_month, y = monthly_approval, color = "Approval")) +
  geom_line(aes(x= year_month, y = inflation * 10, color = "Inflation")) +
  labs(x = "Date",
       y = "Approval/Inflation")
```
The graph above plots the presidential approval rating and the national inflation rate from January 1940 to October 2023. The approval rating is reflected by the scale on the y-axis and ranges from 0-100. The inflation rate does not follow the same scale; it is adjusted for better visualization and vertical comparison with the corresponding approval rating.

Even though it is certain that there does not exist a causal relationship between approval rating and inflation rate, one can observe from the graph and hypothesize that these two variables exhibit an inverse correlation: on average, approval rating moves in the opposite direction of inflation.

There are several spikes in the graph that may elucidate the relationship we described above. For example, in the years post-WW2 and near 1950, there was a significant spike in inflation rate. In the meantime, the president at the time, Truman, suffered a huge blow at public approval. Another spike occurred in the mid-late 1970s, when inflation rate skyrocketed. This was perhaps due to the 1970s oil embargo that had a tremendous impact on the American economy. With increasing inflation rate, the approval rating also went down dramatically. One can also observe that during the periods 1955-1968, 1992-2008, 2010-2020, the inflation rate and approval rate seem to be a vertical mirror of one another, which could also suggest an inverse correlation between these two variables.

```{r}
fit <- lm(monthly_approval ~ inflation, data = approval2000)
fit
summary(fit)
```

